The Unicode Cookbook \linebreak for Linguists

### 2018-06-06 book/chapters/pitfalls.tex
deals with paperhivers through chapter 2
through page 21

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; on the contrary, we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been taken differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected -- often unknowingly -- a \textsc{pitfall}.

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently than expected, we think it is important not
to dismiss Unicode as wrong or deficient, because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium has a more
wide-ranging view of matters and often examines important practical use-cases
that from a linguistic point of view are normally not considered. Our general
guideline for dealing with the Unicode Standard is to accept it as it is, and
not to battle windmills. Alternatively, of course, it is possible to actively
engage in the development of the standard itself, an effort that is highly
appreciated by the Unicode Consortium.

@@

A central principle of Unicode is the distinction between character and glyph. A
character is the abstract notion of a symbol in a writing system, while a glyph
is the concrete drawing of such a symbol. In practice, there is a complex
interaction between characters and glyphs. A single Unicode character may of
course be rendered as a single glyph. However, a character may also be a piece
of a glyph, or vice-versa. Actually, all possible relations between glyphs and
characters are attested.

@@

Second, a single character may be rendered as a sequence of multiple glyphs. For
example, in Tamil one Unicode character may result in a combination of a
consonant and vowel, which are rendered as two adjacent glyphs by fonts that
supports Tamil, e.g.\ \textsc{tamil letter au} at \uni{0B94} represents a single 
character <\tamil{ஔ}>, composed of two glyphs <\tamil{ஓ}> and <\tamil{ன}>. Perhaps confusingly, 
in the Unicode Standard there are also two individual characters, 
i.e.\ \textsc{tamil letter oo} at \uni{0B93} and 
\uni{0BA9} \textsc{tamil letter nnna}, each of which is a glyph. Another example is 
Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <\sinhala{ෛ}>, which is 
visually two glyphs, each represented by \textsc{sinhala vowel sign kombuva} 
at \uni{0DD9} <\sinhala{ ෙ}>.

@@

Further, the rendering of a glyph is dependent on the font being used. For
example, the Unicode character \textsc{latin small letter g} appears as <g> and
<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
because their typefaces are designed differently. Furthermore, font face may
change the visual appearance of a character, for example Times New Roman
two-story <{\fontspec{Times New Roman}a}> changes to a single-story glyph in italics
<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).

In sum, character-to-glyph mappings are complex technological issues that the
Unicode Consortium has had to address in the development of the Unicode
Standard, but for the lay user they can be utterly confusing because visual
rendering does not (necessarily) indicate logical encoding.

@@

The Unicode Standard is a character encoding system, and not a writing system
encoding system. This most forcefully becomes clear with the notion of grapheme.
From a linguistic point of view, graphemes are the basic building blocks of a
writing system (see Section~\ref{linguistic-terminology}). It is extremely
common, up to the point of being universally attested, that writing systems use
combinations of multiple symbols as a single grapheme, like <sch>, <th> or <ei>.
There is no possibility to encode such complex graphemes using the Unicode Standard.

The Unicode Standard deals with complex graphemes only inasmuch they consist of
base characters with diacritics (see
Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
clusters}. Complex graphemes consisting of multiple base characters,
like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
Chapter~\ref{the-unicode-approach}).

Inspecting the Unicode Standard, there appear to be special Unicode characters
to glue together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both code points actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called \textit{collation}
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures.

To solve the issue of tailored grapheme clusters, Unicode offers some assistance
in the form of the Unicode Locales.\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
However, in the practice of
linguistic research, this is not a real solution. For that reason we propose to
use orthography profiles (see Chapter~\ref{orthography-profiles}). Basically,
both orthography profiles and locale descriptions offer a way to specify
tailored grapheme clusters. For example, for English one could specify that <sh>
is such a cluster. Consequently, this sequence of characters is then always
interpreted as a complex grapheme. For cases in which this is not the right
decision, like in the English word \textit{mishap}, the \textsc{zero width
joiner} at \uni{200D} has to be entered between <s> and <h>.

@@

The central problem behind the missing glyphs is that designing actual glyphs
includes a lot of different considerations and it is a time-consuming process.
Many traditional expectations of how specific characters should look like have
to be taken into account when designing glyphs. Those expectations are often not
well documented, and it is mostly up to the knowledge and experience of the font
designer to try and conform to them as good as possible. Therefore, most
designers produce fonts only including glyphs for certain parts of the Unicode
Standard, namely for those characters they feel comfortable with. At the same
time, the number of characters defined by the Unicode Standard is growing with
each new version, so it is neigh impossible for any designer to produce glyphs
for all characters. The result of this is that, almost necessarily, each font
only includes glyphs for a subset of the characters in the Unicode Standard.

The simple solution to missing glyphs is thus to install additional fonts
providing additional glyphs. For the more exotic characters there is often not
much choice. There are a few particularly large fonts that might be considered.
First, there is the \textsc{Everson Mono} font made by Michael Everson, which
currently includes 9,756 different glyphs (not including Chinese) updated up to
Unicode 7.0.\footnote{Everson Mono is available as shareware at
\url{http://www.evertype.com/emono/}.} Already a bit older is the \textsc{Titus
Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes
10,044 different glyphs (not including Chinese), but not including newer
characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available
at \url{http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp}.}

>>>>

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; rather we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been made differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected -- often unknowingly -- a \textsc{pitfall}.

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently than expected, we think it is important not
to dismiss Unicode as wrong or deficient, because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium has a 
wide-ranging view of matters and often examines important practical use-cases
that are not normally considered from a linguistic point of view. Our general
guideline for dealing with the Unicode Standard is to accept it as it is, and
not to tilt at windmills. Alternatively, of course, it is possible to actively
engage in the development of the standard itself, an effort that is highly
appreciated by the Unicode Consortium.

@@

A central principle of Unicode is the distinction between character and glyph. A
character is the abstract notion of a symbol in a writing system, while a glyph
is the visual representation of such a symbol. In practice, there is a complex
interaction between characters and glyphs. A single Unicode character may of
course be rendered as a single glyph. However, a character may also be a piece
of a glyph, or vice-versa. Actually, all possible relations between glyphs and
characters are attested.

@@

Second, a single character may be rendered as a sequence of multiple glyphs. For
example, in Tamil one Unicode character may result in a combination of a
consonant and vowel, which are rendered as two adjacent glyphs by fonts that
support Tamil, e.g.\ \textsc{tamil letter au} at \uni{0B94} represents a single 
character <\tamil{ஔ}>, composed of two glyphs <\tamil{ஓ}> and <\tamil{ன}>. Perhaps confusingly, 
in the Unicode Standard there are also two individual characters, 
i.e.\ \textsc{tamil letter oo} at \uni{0B93} and 
\uni{0BA9} \textsc{tamil letter nnna}, each of which is a glyph. Another example is 
Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <\sinhala{ෛ}>, which is 
visually two glyphs, each represented by \textsc{sinhala vowel sign kombuva} 
at \uni{0DD9} <\sinhala{ ෙ}>.

@@

Further, the rendering of a glyph is dependent on the font being used. For
example, the Unicode character \textsc{latin small letter g} appears as <g> and
<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
because their typefaces are designed differently. Furthermore, the font face may
change the visual appearance of a character, for example Times New Roman
two-story <{\fontspec{Times New Roman}a}> changes to a single-story glyph in italics
<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).

In sum, character-to-glyph mappings are complex technical issues that the
Unicode Consortium has had to address in the development of the Unicode
Standard. However, they can can be utterly confusing for the lay user because visual
rendering does not (necessarily) indicate logical encoding.

@@

The Unicode Standard encodes characters. This becomes most clear with the notion of grapheme.
From a linguistic point of view, graphemes are the basic building blocks of a
writing system (see Section~\ref{linguistic-terminology}). It is extremely
common for writing systems to use 
combinations of multiple symbols (or letters) as a single grapheme, such as <sch>, <th> or <ei>.
There is no way to encode such complex graphemes using the Unicode Standard.

The Unicode Standard deals with complex graphemes only inasmuch as they consist of
base characters with diacritics (see
Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
notion of diacritic). The Unicode Standard calls such combinations \textit{grapheme
clusters}. Complex graphemes consisting of multiple base characters,
like <sch>, are called \textit{tailored grapheme clusters} (see
Chapter~\ref{the-unicode-approach}).

There are special Unicode characters that 
glue together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both code points actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called \textit{collation}
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures.

To solve the issue of tailored grapheme clusters, Unicode offers some assistance
in the form of Unicode Locales.\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
However, in the practice of
linguistic research, this is not a real solution. To address this issue, we propose to
use orthography profiles (see Chapter~\ref{orthography-profiles}). Basically,
both orthography profiles and Unicode Locales offer a way to specify
tailored grapheme clusters. For example, for English one could specify that <sh>
is such a cluster. Consequently, this sequence of characters is then always
interpreted as a complex grapheme. For cases in which this is not the right
decision, like in the English word \textit{mishap}, the \textsc{zero width
joiner} at \uni{200D} has to be entered between <s> and <h>.

@@

The central problem behind the missing glyphs is that designing actual glyphs
includes many different considerations and it is a time-consuming process. 
Many traditional expectations of what specific characters should look like have
to be taken into account when designing glyphs. Those expectations are often not
well documented, and it is mostly up to the knowledge and experience of the font
designer to try and conform to them as well as possible. Therefore, most
designers produce fonts that only include glyphs for certain parts of the Unicode
Standard. At the same
time, the number of characters defined by the Unicode Standard is growing with
each new version, so it is nigh impossible for any designer to produce glyphs
for every character. The result of this is that, out of necessity, each font
only includes glyphs for a subset of the characters in the Unicode Standard.

A simple solution to missing glyphs is to install additional fonts
providing additional glyphs. For the more exotic characters there is often not
much choice. There are a few particularly large fonts that might be considered.
First, there is the \textsc{Everson Mono} font made by Michael Everson, which
currently includes 9,756 different glyphs (not including Chinese) updated up to
Unicode 7.0.\footnote{Everson Mono is available as shareware at
\url{http://www.evertype.com/emono/}.} Already a bit older is the \textsc{Titus
Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes
10,044 different glyphs (not including Chinese), but not including newer
characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available
at \url{http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp}.}
